---
title: "lab_week9"
author: "Anish"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc : true 
    theme : united 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
library(dplyr)
library(ggplot2)
```


```{r cars}
teams = read.csv("teams.csv")
tail(teams)
str(teams)

```

## Q1

```{r pressure, echo=FALSE}
myteams <-  select(teams, G, W, L, R, RA, teamID, yearID, lgID)
myteams <- filter(myteams, yearID > 2000)
```

```{r}
myteams$RD <- with(myteams, R - RA)
myteams$Wpct <- with(myteams, W / (W + L))
```

```{r}
 boxplot(Wpct ~ lgID, data=myteams)
```
## Q1
```{r}
ggplot(myteams, aes(y = Wpct)) + geom_boxplot() + facet_wrap(~lgID) + labs(y = "Winning Percentage", x = "American League vs National League")

```
```{r}
WP.AL <- subset(myteams,lgID =="AL")$Wpct
WP.NL <- subset(myteams,lgID =="NL")$Wpct
qqnorm(WP.AL)
qqnorm(WP.NL)
t.test(WP.AL,WP.NL)

```
## Q2

```{r}
twoteams <- myteams %>% filter(teamID == "CLE" | teamID == "BOS")
twoteams <- select(twoteams, teamID, Wpct)
ttest <- t.test(Wpct~teamID,twoteams) 
ttest
```
The p value is lesser than .05, so there is a significant difference between the win percentages of the Boston Red Sox and Cleveland Indians.


```{r}
plot(myteams$RD, myteams$Wpct, 
     xlab = "Run Differential", 
     ylab ="Winning Percentage")
```
```{r}
ggplot(myteams)+ geom_point(aes(x = RD, y=Wpct)) + geom_smooth(aes(x = RD, y = Wpct), method = 'lm')
ggplot(myteams, aes(x = RD, y=Wpct))+ geom_point()+ geom_smooth(method = 'lm')
```

## Q3
```{r}
ggplot(myteams %>% filter(lgID == "AL"))+ geom_point(aes(x = RD, y=Wpct)) + geom_smooth(aes(x = RD, y = Wpct), method = 'lm') + labs (x = "Run Differtential", y = "Win percentage")
```
```{r}
linfit <- lm(Wpct ~ RD, data = myteams)
 linfit$coefficients
 par(mfrow=c(2,2))
 plot(linfit)
```
```{r}
plot(myteams$RD, myteams$Wpct, 
     xlab = "Run Differential", 
     ylab = "Winning Percentage",)
abline(linfit$coefficients[1],linfit$coefficients[2])

```
##Q4
```{r}
ALdata <- filter(myteams, lgID == "AL")
linfit <- lm(Wpct ~ RD, data = ALdata)
linfit$coefficients
par(mfrow=c(2,2))
plot(linfit)
plot(myteams$RD, myteams$Wpct, 
    xlab = "Run Differential", 
    ylab = "Winning Percentage",)
abline(linfit$coefficients[1],linfit$coefficients[2])
```



## Q5
What is the difference in the average winning percentage of the New York Yankees and Texas Rangers?
```{r}
twoteams_2 <- myteams %>% filter(teamID == "TEX" | teamID == "NYA")
twoteams_2 <- select(twoteams_2, teamID, Wpct)
ttest_2 <- t.test(Wpct~teamID,twoteams_2) 
ttest_2

```
The p-value is 5.217 * 10^-5, which is lesser than .05. This means that there is a statistically significant difference in the the average winning percentage of the Yankees and Rangers. 0 also isn't part of the 95% confidence interval, thus making the difference more evident. 
